---
title: "Spatial Autocorrelation in Machine Learning."
output: html_document
---

<!--
IMPORTED - not part of the original published archive.

  source repo   Glenn-Prosopis-ML
  source path   spatial-autocorrelation.Rmd
  ref           master @ 6d9def5
  sha256        3295dfc503a8f2c2109b4d7ac063971c3b88f1ddcd7d9abdf31d45a264dc554d

  why           The ONLY place in any repository where spcv_block appears.
                Compares random / block / kmeans / disc resampling and calls
                autoplot(show_blocks = TRUE) - exactly the figure Reviewer 1
                asked for at L239-242. Likely seeded the manuscript's
                "spatial block cross-validation" wording, though it runs on
                Bokspits_1 only, on a 5x-aggregated cube, with an untuned
                randomForest at 6 folds x 3 repeats - NOT the configuration
                whose accuracies the manuscript reports.

Content below is VERBATIM. It expects to run from the Glenn-Prosopis-ML
project root (data_in/, data_out/, R/). See refactor-findings.md.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Why we should be careful when evaluating model performance...

This notebook provides a brief example of why we should take care when building
machine learning (ML) models with a spatial dimension. And, specifically, why
we must be careful when we evalutate model performance.

*Do not use spatial location as a predictor variable (unless you
have a very good reason). This will bring out any spatial pattern in the predictions
and will be reflected in predictions in novel/untrained areas.*


Here we use Glenn's Prosopis dataset at the Boxspits 1 site. It is a great example
of Land Use Classification and is a great illustration of the potential issue.

firstly, let's read in the raster data cube -  note there are lots of bands which
we will use as predictors.
```{r read-n-rescale}
invisible(lapply(list.files("R", full.names = TRUE, recursive = TRUE), source))

input_ras_cube <- rast("data_out/Bokspits_1/Bokspits_1_stack.tif") |> 
  aggregate(5,  filename="data_out/Bokspits_1/Bokspits_1_stack_agg h.tif", overwrite=TRUE)  # reduce resolution for the example.

pnts <- vect("data_in/Bokspits_1/Bokspits_1_Field_data_points_All_b30.shp")

terra::plotRGB(input_ras_cube[[c(3,2,1)]], scale=0.6, smooth=TRUE, axes=FALSE)
points(centroids(pnts), col="cyan", cex=0.5)

plot(input_ras_cube, col=grDevices::hcl.colors(50, palette = "Sunset"))

```

Much of the next bit for generating the training dataframe is hidden in the 
`build_ml_df` function but don't worry. Essentially, Glenn labelled the ground 
cover at numerous locations and we then covert the are of these locals (30 cm 
disk) to raster and extract coincident values from the raster cube.

Let's also make some predictions using a random forest model - as you can see, 
there appears to be good separation of different land cover features...

```{r build-ml-df-predict}

v <- build_ml_df(cube = input_ras_cube,
                 site_name = "Bokspits_1",
                 df_type="grid")

task <- build_task(v, site_name="Bokspits_1")

head(task$data())


rf_lrnr <- lrn("classif.randomForest")

rf_lrnr$train(task)

preds <- terra::predict(input_ras_cube, rf_lrnr, na.rm=TRUE, cores=4)

plot(preds, col=rev(brewer.pal(6, name="Dark2")))

```

But how accurate is the model?? The way we evaluate this with ML models is to 
use resampling. i.e, split the data into training and testing sets, train the 
model with the train portion and then test on the reduced portion. Ideally, we
do this multiple times, in this case we use 6 folds, therefore the data is trained
and tested 6 times (repeated CV is a good idea for a more thorough evaluation).


This is the standard approach for cross validation - simple random splitting of 
the data. This is fine, and even preferred, if your data have no underlying autocorrelation 
structure such as time or space. However, let's think about why this might be a
problem... If things closer to each other are more likely to be similar, then 
training the model and testing on neighbours is more likely to give an optimistic 
evaluation score. When we apply this model to new unseen areas, we are very likely
overestimating model accuracy.

```{r random-cv}
rand_cv <- mlr3::rsmp("cv", folds=6)

autoplot(rand_cv, task=task, fold_id=1:4)
```


So what can we do? Well, we can explicitly consider the autocorrelation strucure
in the resampling approach. In this case we are just thinking about spatial 
autocorrelation but the same applies for temporal - we shouldn't test and train 
on neighbouring or repeated points from differing time periods.

The simplest approach is the Block CV method. overlay a grid and use this to 
split the dataset.


```{r}
block_cv <- mlr3::rsmp("spcv_block", folds = 6, range = 100)
autoplot(block_cv, task=task, fold_id=1:4,show_blocks = TRUE)

```

Many other methods exist also, for example:

Spatial clustering using kmeans...

```{r}
spcv <- mlr3::rsmp("spcv_coords", folds = 6)
autoplot(spcv, task=task, fold_id=1:4)

```


spatial clustering with buffered/disk seperation...

```{r}
spcv_buff <- mlr3::rsmp("spcv_disc", folds = 6, radius=50, buffer=50)
autoplot(spcv_buff, task=task, fold_id=1:4)

```

Let's look at what this means for model accuracy... quite a big difference. In fact,
depending on the data, this can be much more extreme. 
Note that the random CV is more optimistic than the other approaches. It's a 
good idea to put some thought into your resampling regime - this will be dependent 
on a number of factors including how much training data you have and how it is
distributed spatially.

```{r}

resample_results <- function(res_type, res_plan){
  rs <- ml_resample(.task = task,.learner = rf_lrnr,
                     spcv_plan = res_plan, .folds = 6, .repeats = 3)
  message(res_type)
  print(rs$aggregate(c(msr("classif.acc"),
                    msr("classif.ce"))))
  tibble({{res_type}} := rs$score(msr("classif.acc"))$classif.acc)
}


name_list <- c("Random CV", "Block CV", "Coordinate-based k-means", "Spatial 'disc' resampling")
rs_list <- c(rand_cv, block_cv, spcv, spcv_buff)

all_resamps <- purrr::map2(name_list, rs_list, ~resample_results(.x, .y)) |> 
  purrr::list_cbind() 

all_resamps |> 
  tidyr::pivot_longer(everything())|> 
  dplyr::mutate(name = forcats::fct_relevel(name, name_list)) |> 
  ggplot() +
  aes(x = name, y = value, fill=name) +
  geom_violin(trim = TRUE, alpha=0.8) +
  geom_boxplot(fill=NA, width=0.2)+
  geom_jitter(height = 0, width = 0.1) +
  scale_fill_brewer(palette="Set2")+
  theme_linedraw() +
  guides(fill="none") +
  labs(x="Resampling Type", y= "Rerported Accuracy")
  
```
