---
title: "Detecting sparse biological invasions through multi-scale remote sensing: a drone–satellite framework for mapping Neltuma cover, with calibrated uncertainty, in dryland ecosystems"
author: "Slade et al. (anonymised) — refactor-3.0 draft"
date: today
format:
  html:
    embed-resources: true
    css: pipeline.css
    toc: true
  docx:
    toc: false
execute:
  echo: false
  warning: false
---

<!--
THIS DOCUMENT IS A PIPELINE PRODUCT. Every number is an inline expression
reading `paper_values`; figures and tables are targets outputs. The prose is
the submitted manuscript where its claims still hold under the refactored
analysis, and rewritten where they do not. Each rewritten claim is listed for
the co-authors in paper/changes-from-submitted.md with the finding it rests
on. Values are highlighted in green in the HTML render so a reader can see
what the pipeline controls; the highlighting is CSS only and absent in docx.
Render via targets: the `paper` target (tarchetypes::tar_quarto). The
Supplementary Information is paper/supplement.qmd (the `supplement` target).
-->

```{r setup}
#| include: false
# Resolve everything from the project root so the document renders identically
# under tar_make (cwd = root) and by hand from paper/ (cwd = paper).
options(knitr.kable.NA = "–")
root  <- if (file.exists("_targets.R")) normalizePath(".") else normalizePath("..")
store <- file.path(root, basename(Sys.getenv("NELTUMA_STORE", "_targets")))
# Literal tar_read() calls so tar_quarto registers the dependencies.
values      <- targets::tar_read(paper_values, store = store)
sensor_tab  <- targets::tar_read(sensor_summary, store = store)
phase_tab   <- targets::tar_read(cover_phase_index, store = store)
stacks_tab  <- targets::tar_read(stacks, store = store)
figs <- list(
  fig_study = targets::tar_read(fig_study, store = store),
  fig_maps = targets::tar_read(fig_maps, store = store),
  fig_acc = targets::tar_read(fig_acc, store = store),
  fig_cover = targets::tar_read(fig_cover, store = store),
  fig_sensors = targets::tar_read(fig_sensors, store = store),
  fig_cover_grain = targets::tar_read(fig_cover_grain, store = store),
  fig_cover_area = targets::tar_read(fig_cover_area, store = store),
  fig_cover_coverage = targets::tar_read(fig_cover_coverage, store = store))
fig <- function(name) normalizePath(file.path(root, figs[[name]]))
stacks_tab <- stacks_tab[, c("tag", "n_bands", "bands")]
stacks_tab$bands <- gsub("|", ", ", stacks_tab$bands, fixed = TRUE)
stacks_tab$label <- c("5" = "Drone", "5_CHM" = "Drone + CHM", "5_CHM_NDVI" = "Drone + CHM + NDVI",
                      "5_CHM_ALLVI" = "Drone + CHM + All VI")[stacks_tab$tag]
sensor_label <- c(wv2 = "WorldView-2", planet = "PlanetScope", s2 = "Sentinel-2")
```

::: {.pipeline-legend}
**Draft for co-author review.** [Green]{.substituted} marks a value written by the analysis pipeline. Grey boxes are notes to the co-authors and are not part of the manuscript text.
:::

# Abstract

Biological invasions are transforming dryland ecosystems worldwide, yet detecting sparse invasive plants remains a fundamental Earth observation challenge because ecological targets often occur below the grain of operational sensors. Here, we develop a hierarchical drone–satellite remote sensing framework to detect and map invasive *Neltuma* (formerly *Prosopis*) across heterogeneous dryland ecosystems, integrating UAV imagery, field observations and optical satellite data within a linked multi-scale machine learning workflow, and we pair every landscape estimate with a calibrated statement of its uncertainty.

Using drone surveys, we achieved overall classification accuracies of [`r values$stack_best_pct`]{.substituted}% under prediction-domain-matched spatial cross-validation and [`r values$avg_neltuma_recall_pct`]{.substituted}% recall for *Neltuma*, with 84.7% of independently mapped *Neltuma* plants (n = 184) detected. Fine-resolution observations alone are, however, difficult to scale, and a discrete satellite classification proved an unreliable route to landscape extent: evaluated honestly against the drone maps, WorldView-2 (1.6 m) recovered [`r values$px_wv2_recall_pct`]{.substituted}% of drone-mapped *Neltuma* pixels at [`r values$px_wv2_precision_pct`]{.substituted}% precision, while PlanetScope (3 m) and Sentinel-2 (10 m) over-predicted *Neltuma* extent by factors of [`r values$px_planet_ratio`]{.substituted} and [`r values$px_s2_ratio`]{.substituted}. We therefore modelled *Neltuma* as continuous sub-pixel cover, using calibrated drone probabilities as the training target, and attached distribution-free conformal prediction intervals, an explicit area of applicability and a prediction-powered area correction. Within the area over which its intervals hold, WorldView-2 estimated [`r values$cover_wv2_ha`]{.substituted} ha of *Neltuma* cover (95% CI [`r values$cover_wv2_lo`]{.substituted}–[`r values$cover_wv2_hi`]{.substituted} ha) across the 445 km² study area, with PlanetScope and Sentinel-2 giving concordant totals. Sensor grain governed not detection alone but the extent over which detection could be trusted: at 1.6 m the dense settlement and river-corridor invasion fronts fell within the area of applicability, whereas at 3–10 m they were increasingly flagged as extrapolation. Invasion is overwhelmingly sparse, with [`r values$cover_wv2_expansion_pct`]{.substituted}% of the landscape in an active expansion phase and negligible dense dominance, and cover increases towards settlements and roads.

Beyond species mapping, our results show how integrating fine-resolution ecological observations with satellite Earth observation via a hierarchical drone–satellite framework can help overcome scale mismatch, and how honest spatial validation, conformal uncertainty and an area of applicability turn a map into a decision-ready estimate. The framework has wider relevance for biodiversity surveillance, woody encroachment, degradation monitoring and dryland early-warning systems.

# Keywords

Mesquite, Prosopis, Woody plant encroachment, Rangeland degradation, Ecological monitoring, Environmental intelligence, Conformal prediction, Area of applicability

# 1. Introduction

Rangeland degradation is intensifying across African drylands, threatening livelihoods, biodiversity and ecosystem function through interacting pressures from land-use change, invasive species and climate change. In dryland systems these pressures are often spatially heterogeneous and difficult to monitor, creating major challenges for environmental management and raising broader Earth observation questions about how to detect sparse ecological change across landscapes. In Botswana's rangelands, invasion by exotic plants, particularly the *Neltuma* genus (Figure 1), is a prominent example of this challenge (Dougill et al. 2016). Formerly named *Prosopis* (Hughes et al. 2022) and commonly known in Botswana as Mesquite or Sexanana, *Neltuma* species in southwest Botswana have formed a hybrid swarm comprising four pure lines (*N. chilensis*, *N. juliflora*, *N. velutina* and *N. glandulosa*) and at least nine hybrids and variants. In this study, *Neltuma* refers to all species, variants and hybrids (Muzila et al. 2010).

*Neltuma* invasion is a socially and ecologically consequential form of environmental change in African rangelands. As a globally significant invasive alien genus, *Neltuma* can alter ecosystem structure, compete strongly for water and transform savanna vegetation dynamics (Hussain et al. 2021). In South Africa, native woody species density, richness and diversity decline with increasing *Neltuma* encroachment (Shackleton et al. 2015), while canopy dieback and mortality have been recorded in keystone Kalahari savanna trees (Schachtschneider and February 2013). In southwest Botswana, local communities associate *Neltuma* spread with falling water tables and report broader impacts on health and livelihoods (Mosweu et al. 2013). Yet these invasions are complex, with reported benefits including dune stabilisation, windbreaks, livestock fodder and fuelwood, complicating management responses (Mosweu et al. 2013).

Effective management of *Neltuma* depends on spatial information that has largely been lacking. Botswana's strategy of control by utilisation recognises both the ecological threat and socio-economic complexity of invasion, and the Integrated Mesquite (Prosopis Species) Management Strategy for Botswana (2022) calls for context-specific responses in affected communal areas. However, planning, prioritisation and monitoring require reliable spatial information on invasion extent, density and progression at both local and landscape scales. Such information has been difficult to obtain using accessible mapping techniques.

Detecting sparse invasive plants across heterogeneous drylands remains a persistent scale-mismatch problem in Earth observation. Measuring *Neltuma* distribution in southwest Botswana is particularly challenging because the invasion often occurs at low fractional cover, spans diverse growth forms and phenological stages, and is embedded within heterogeneous vegetation mosaics. Effective monitoring therefore requires methods capable of detecting individual plants during early incursion while also capturing low-density populations across broader landscapes (Müllerová et al. 2023). Satellite remote sensing has proven effective for mapping some invasive plants (Robinson et al. 2016; Rajah et al. 2019; Rebelo et al. 2021a), and *Neltuma* has been mapped using WorldView-2, Sentinel-2 and Landsat imagery (Adam, Mureriwa, and Newete 2017; Berg and Catharina 2013; Eckert et al. 2025; Page et al. 2022), but satellite approaches remain constrained by spatial and spectral resolution when ecological targets are sparse or patchy (McMillan et al. 2023). This can limit sensitivity precisely where early detection is most important.

Two further problems compound the scale mismatch, and both concern how confident a map is rather than how it is made. First, the accuracy of a species map depends on how it is evaluated: spatially structured training data produce optimistic accuracy under random cross-validation, and even spatial cross-validation returns different numbers depending on how far the held-out data lie from the training data relative to the distances the map must bridge (Roberts et al. 2017; Meyer and Pebesma 2022; Milà et al. 2022; Linnenbrink et al. 2024). Second, a classification assigns each pixel to a single class and reports no per-pixel uncertainty, so a landscape total derived from it carries no interval and cannot say where the model is extrapolating beyond its training domain (Meyer and Pebesma 2021). Distribution-free conformal prediction (Angelopoulos and Bates 2022) and prediction-powered inference (Angelopoulos et al. 2023) offer principled remedies for the second problem, and prediction-domain-adaptive cross-validation (Linnenbrink et al. 2024) for the first, but they have rarely been combined in invasion mapping.

Integrating drone and satellite observations offers new opportunities to bridge scale gaps in ecological monitoring. Drone-based remote sensing can capture ecological structure at resolutions inaccessible to satellites and provides a valuable intermediary between field observations and landscape-scale Earth observation (Alvarez-Taboada, Paredes, and Julián-Pelaz 2017; Baena et al. 2017; Kattenborn et al. 2019; Cunliffe, Brazier, and Anderson 2016). Combined with satellite imagery, drone data create opportunities for multi-scale analytical frameworks that improve interpretation of coarser imagery, strengthen ecological inference and support more operational environmental monitoring. In particular, a drone classification resolves the sub-pixel composition of every coarser satellite pixel it overlaps, which turns the scale mismatch from an obstacle into a training signal: rather than forcing a mixed pixel into one class, the fraction of that pixel occupied by the target can be estimated directly (Frazier and Wang 2011; Shiferaw et al. 2019). Such approaches can also help guide local control and eradication efforts, inform policy and management planning, and enable monitoring of follow-up interventions through repeatable, scalable observations.

This study assessed whether *Neltuma* can be mapped in ways useful to ecologists, land users and managers, with an emphasis on stating how far each product can be trusted. We combined ground observations with remote sensing from drone and satellite sensors to address four research questions: (1) Can drone-derived data reliably detect *Neltuma* at local scales in the Kalahari, and which combinations of data and machine learning algorithms maximise accuracy? (2) Can drone-derived products calibrate satellite estimates of sub-pixel *Neltuma* cover, and with what uncertainty? (3) How does sensor grain constrain the detection of *Neltuma*, and over what part of the landscape can each sensor's estimate be trusted? (4) What insights does remote sensing provide into the invasion dynamics of *Neltuma* in the region?

# 2. Methodology

## 2.1 Study system and research design

We used a hierarchical multi-scale Earth observation design to investigate sparse *Neltuma* invasion, integrating field observations with drone and optical satellite imagery. The study was structured to evaluate plant-scale detection, to use fine-resolution observations as calibration and validation data for satellite-based estimates, to quantify the effects of sensor grain on invasion detection, and to characterise spatial patterns of invasion across the study area. Fine-resolution drone observations provided detailed ecological information for calibrating and evaluating satellite products, while satellite imagery enabled assessment of how detection performance and its trustworthy extent changed as spatial resolution decreased. By linking these observation scales, the design allowed us to examine both the value of fine-resolution data for improving landscape-scale inference and the limitations introduced when sparse invasion patterns are observed at coarser spatial grains.

The study was conducted across approximately 445 km² of Kgalagadi rangelands in southwest Botswana surrounding the villages of Bokspits, Struizendam and Vaalhoek, bounded to the south by the Molopo and to the west by the Nossob ephemeral river valleys, which also mark the Botswana–South Africa border (Figure 1). The region comprises a semi-arid savanna system of deep sandy soils, variable grazing pressure and heterogeneous vegetation structured by NW–SE dune systems, dune valleys and dry river corridors. Vegetation is dominated by Kalahari Duneveld shrub-grassland, with dune crests supporting *Stipagrostis amabilis* and scattered trees such as *Vachellia haematoxylon* and *Boscia albitrunca*, while dune valleys contain grassland and low-density bush savanna with species including *Schmidtia kalahariensis*, *Stipagrostis uniplumis*, *Eragrostis lehmanniana* and *Aristida meridionalis*, interspersed with denser shrub patches of *Rhigozum trichotomum* and *Senegalia mellifera*. Dry river valleys support more open tree savanna including *V. erioloba* (Van Rooyen and Van Rooyen 2019). Communal pastoralism is the dominant land use, with continuous grazing by livestock shaping vegetation structure and disturbance regimes (Kgosikoma and Batisani 2014). Within this dynamic dryland mosaic, *Neltuma*, introduced during sand dune stabilisation near Bokspits in 1980 (van der Poel and Timberlake 1980), has spread from initial planting sites into settlements and surrounding savanna (Mosweu et al. 2013), occurring across a continuum from isolated individuals and low-density populations to dense established stands, making it a suitable system for evaluating scale-dependent detection.

```{r fig1-panel, out.width="60%"}
knitr::include_graphics(fig("fig_study"))
```

::: {.pipeline-note}
Pipeline analogue of Figure 1D: boundary and the seven sites. Panels A–C (location), E–F (photographs) and the Sentinel-2 basemap are authoring items to be restored from the submitted figure.
:::

Figure 1. Multi-scale remote sensing of sparse *Neltuma* invasion is situated within a heterogeneous Kalahari dryland landscape. (A–C) Location of the study area within Africa, Botswana and the Kgalagadi District. (D) The 445 km² study area surrounding Bokspits, Struizendam and Vaalhoek, showing Kalahari dune systems, settlements, drone survey sites and study area boundaries over Sentinel-2 imagery. (E) Representative Kalahari Duneveld vegetation illustrating the heterogeneous shrub–grassland mosaic in which sparse *Neltuma* occurs. (F) Dense *Neltuma* invasion near settlements and transport corridors.

## 2.2 Ground observations

Ground observations provided reference data for model training, independent validation and ecological interpretation across the multi-scale framework. Field surveys were designed to capture variation in *Neltuma* occurrence, density, growth form and associated land-cover classes across the invasion gradient, including isolated individuals, low-density populations and dense established stands, with emphasis on sparse occurrences that present the greatest detection challenge.

Field observations were collected between 1 September and 2 October 2022 (Table S2), during the late dry season when predominantly evergreen *Neltuma* contrasted strongly with largely quiescent or marcescent vegetation, reducing phenological mismatch with remotely sensed data. Surveys spanned all seven drone survey areas and recorded representative observations for the principal land-cover and vegetation classes used in classification (Tables S3 and S4).

Training and validation points were digitised in the field using QField v2.6.2 on GNSS-enabled tablets displaying orthomosaics generated from drone surveys processed on site in Pix4DMapper v4.8.0 (Pix4Dmapper 2022). Sampling points were selected so an 80 cm radius contained only a single land-cover class, with a target of 20 observations per class in each drone survey area; [`r values$n_field_total`]{.substituted} observations were retained across the seven areas, and the target was not met for [`r values$n_site_class_below20`]{.substituted} of [`r values$n_site_class_combos`]{.substituted} site–class combinations, the smallest being *V. erioloba* at Struizendam 4 (n = [`r values$min_class_n`]{.substituted}; Table S4). Spectral values associated with each field point were extracted using the R package exactextractr (Baston 2023), sampling pixels within 30 cm of point centres to calculate weighted averages of reflectance values; each field point represented approximately 80–105 image pixels depending on terrain and flight height. Independent plant-scale validation surveys were conducted in two 100 × 200 m plots within drone survey areas, where the central location and maximum height of every *Neltuma* individual (n = 184) were recorded alongside representative observations of other vegetation classes and bare ground. These data were used to assess plant-level detection performance across growth forms and phenological states and to complement pixel-based accuracy assessment. Additional field observations were used to interpret ecological associations relevant to invasion dynamics, including frequent co-occurrence of *Neltuma* beneath native woody canopies. Together, these observations provided the ecological reference data underpinning classification development, validation and interpretation across scales.

## 2.3 Drone observations and image processing

Drone observations provided the fine-resolution ecological observation layer linking field measurements to landscape-scale satellite analysis. UAV surveys were designed to resolve individual plants at spatial grains beyond the capability of satellite sensors while generating ecological products for cross-scale calibration.

Drone surveys were conducted across seven sites spanning gradients in *Neltuma* density, geomorphology and vegetation structure near Bokspits and Struizendam (Figure 1D; Table S2), with site selection balancing ecological representativeness, community preferences and Botswana Civil Aviation Authority regulations. Surveys were undertaken between 1 September and 2 October 2022 during the late dry season to maximise spectral contrast between predominantly evergreen *Neltuma* and co-occurring species such as *R. trichotomum* and *S. mellifera* in largely quiescent or marcescent states. Flights were conducted under low-wind, clear-sky conditions to minimise artefacts in photogrammetric reconstruction, given evidence that canopy heights derived from drone photogrammetry can be sensitive to wind speed (Slade et al. 2025).

Imagery was acquired using a DJI Phantom 4 Multispectral (P4MS) with a six-camera array (Table S1), paired with a DJI D-RTK2 High Precision GNSS Mobile Station for positional control. Flight plans were implemented in Ground Station Pro using nadir acquisition at 90–100 m above ground level, 80% forward and side overlap, and a 2.1 s image interval, yielding 5.8–6.7 cm ground sampling distance.

Radiometric calibration followed empirical line methods using a 50% Zenith Lite Diffuse Reflectance Target, with calibration images collected pre- and post-flight and processed following established multispectral workflows (Assmann et al. 2018; Smith and Milton 1999). Multispectral imagery was processed in Pix4DMapper v4.8.0 to generate calibrated orthomosaics for each spectral band. RGB imagery was processed using structure-from-motion and multi-view stereopsis to generate dense 3D point clouds, from which Digital Surface Models and Digital Terrain Models were used to derive Canopy Height Models (CHM), later resampled to harmonise with multispectral data for classification analyses.

Drone-derived predictor sets included calibrated spectral bands, vegetation indices and the CHM, combined into four predictor stacks (Table 1) and evaluated in combination with alternative machine-learning algorithms to identify configurations maximising classification performance. Drone imagery was used both to assess local-scale detectability of sparse invasions and to generate the calibrated *Neltuma* probability surfaces that serve as the sub-pixel cover target for the satellite models (Section 2.7). Independent plant-level validation was used to assess detection performance before integrating these products into landscape-scale analyses.

Table 1. The four drone predictor stacks (treatments) compared in Section 3.1. The CHM is the photogrammetric canopy height model; the vegetation indices are defined in Table S6.

```{r table1-stacks}
knitr::kable(stacks_tab[, c("label", "tag", "n_bands", "bands")], row.names = FALSE,
             col.names = c("Treatment", "Stack id", "Bands", "Predictors"))
```

## 2.4 Satellite observations and scale-mismatch analysis

Satellite observations were used to evaluate detection of sparse *Neltuma* invasions across contrasting sensor grains using WorldView-2 (1.6 m), PlanetScope (3 m) and Sentinel-2 (10 m), representing fine-, intermediate- and moderate-resolution optical Earth observation systems. Imagery was selected to align as closely as possible with field and drone surveys while minimising phenological mismatch (Table S5), providing a structured basis for assessing how sensor grain influences detectability. Pre-processing included sensor-specific radiometric and atmospheric correction, after which predictor layers were assembled from spectral bands and vegetation indices (NDVI, SAVI, MSAVI2 and MTVI2; Table S6), selected to capture complementary variation in vegetation greenness and canopy signal under sparse vegetation and exposed-soil conditions (Rajah et al. 2019; Rebelo et al. 2021b; Francis et al. 2024; Slade et al. 2023).

Because sparse training targets make classifications sensitive to positional error, satellite imagery was co-registered to drone data using maximum mutual information matching implemented through the coregisterImages function in RStoolbox (v0.3.0) (Leutner et al. 2022), correcting residual offsets among sensors (Rufin et al. 2021).

The drone classifications were linked to each satellite grid in two ways. For the discrete classification (Section 2.5), fractional cover of each class was extracted from the drone classifications to grids matching each satellite pixel footprint using exactextractr (Baston 2023), and pixels whose dominant class exceeded a sensor-specific purity threshold (95%, 85% and 65% for WorldView-2, PlanetScope and Sentinel-2) were used as training labels, balanced to equal class sizes (Table S7). Because heterogeneous vegetation mosaics rarely produce uniform coarse pixels, this route discards the overwhelming majority of the cross-scale information (Figure 5). For the sub-pixel cover product (Section 2.7) no pixel was discarded: every satellite pixel overlapping a drone survey area became a training cell whose target is the drone-derived *Neltuma* cover fraction, giving [`r values$cover_wv2_ntrain`]{.substituted}, [`r values$cover_planet_ntrain`]{.substituted} and [`r values$cover_s2_ntrain`]{.substituted} labelled cells for WorldView-2, PlanetScope and Sentinel-2 respectively from the same seven surveys.

Scale mismatch was assessed by comparing detection performance among sensors, first for the discrete classification against the drone maps and then for the sub-pixel cover product, including the extent of each sensor's area of applicability. These comparisons formed the basis for evaluating sensor-grain constraints and the contribution of multi-scale integration to landscape-scale invasion mapping.

## 2.5 Hierarchical multi-scale classification framework

We implemented a hierarchical multi-scale classification framework to integrate field, drone and satellite observations for detecting and mapping *Neltuma* across nested spatial scales (Figure 2). The framework comprised two linked machine-learning pipelines: a drone workflow generating fine-resolution classifications and calibrated *Neltuma* probability surfaces, and a satellite workflow evaluating sensor performance and estimating sub-pixel *Neltuma* cover with its uncertainty. The whole analysis is a reproducible pipeline (targets, R; Open Research), and every number in this paper is written by it.

Drone classifications combined calibrated multispectral bands, vegetation indices and CHM predictors with field-derived training data. To identify the best-performing classifier we compared five tuned learners, a penalised multinomial regression (glmnet), a support vector machine with linear or radial kernel (SVM), a random forest (RF), and two gradient-boosted tree ensembles (XGBoost and LightGBM), together with an untuned random forest as a baseline, across the four predictor stacks of Table 1. Supervised classification was implemented in R using the mlr3 framework (Lang et al. 2019). Hyperparameters were tuned once per task by model-based (Bayesian) optimisation (mlr3mbo, [`r values$tune_evals`]{.substituted} evaluations) under an inner spatial cross-validation, and the chosen configuration was then evaluated under the outer spatial cross-validation described in Section 2.6. Because learner rankings proved to lie within the fold-to-fold noise (Section 3.1) and because a single tuned model's map of a class occupying about 1% of the landscape is one draw from a wide distribution of equally accurate maps, the landscape products are not the map of a winning learner: each unit's class probabilities are the equal-weight average of the five tuned learners refitted on all of that unit's training data. An equal-weight average has no free parameters to overfit, reduces the variance of sparse-class areas, and composes cleanly with the conformal calibration below; the per-learner surfaces are retained as a sensitivity analysis (Table S14).

An analogous workflow was applied to WorldView-2, PlanetScope and Sentinel-2 for the discrete classification, using three training sources: field observations aligned to each sensor (training points buffered by 0.8 m for WorldView-2), the purity-thresholded drone labels re-derived from our own drone classifications (the primary source), and, as a reproduction check, the purity extraction archived with the original analysis. The same tuning and evaluation framework was applied across sensors.

Model performance was assessed from cross-validation using confusion-matrix-derived overall and class-level accuracies, with particular emphasis on recall of *Neltuma*, and, for the satellites, by pixel-level comparison against the drone classifications within the survey areas. This framework propagated observations across scales, evaluated sensor-grain effects and tested whether centimetre-scale ecological information strengthened landscape-scale inference.

## 2.6 Spatial validation and uncertainty assessment

Classification robustness was assessed with prediction-domain-adaptive spatial cross-validation, in which folds are constructed so that the distribution of nearest-neighbour distances between held-out and training observations matches the distribution between the map's prediction domain and the training observations (k-fold nearest-neighbour distance matching, kNNDM; Linnenbrink et al. 2024; CAST, Meyer et al. 2024). The mismatch between the two distributions is summarised by a Wasserstein statistic *W*, which we report for every task (Table S12) as the quantitative statement of how much extrapolation each map demands. The design separates two legitimately different questions. For the drone classifications the map is the surveyed site, so the prediction domain is the site boundary and the folds are near-random ([`r values$k_unit`]{.substituted} folds, *W* = [`r values$drone_W_min_m`]{.substituted}–[`r values$drone_W_max_m`]{.substituted} m). For the satellite classifications the map is the 445 km² study area, so every fold holds out whole survey areas ([`r values$k_aoi`]{.substituted} folds, *W* = [`r values$sat_W_min_km`]{.substituted}–[`r values$sat_W_max_km`]{.substituted} km): a satellite accuracy reported here is the accuracy of predicting an unsurveyed part of the landscape from the seven surveyed areas. Each design was repeated [`r values$n_repeats`]{.substituted} times with independently seeded fold constructions, giving [`r values$n_outer_iterations`]{.substituted} evaluation iterations per drone task and [`r values$n_outer_iterations_sat`]{.substituted} per satellite task, and the same folds were shared by every learner so that their scores are paired. The comparison with conventional random cross-validation, and the dependence of the satellite accuracy on the fold geometry, are reported in Section 3.2.

Per-pixel uncertainty for the discrete classifications was quantified with class-conditional (Mondrian) split conformal prediction (Angelopoulos and Bates 2022), calibrated on the out-of-fold averaged probabilities of the spatial cross-validation. At a nominal coverage of 90% each pixel receives a prediction set of classes that contains the true class with at least that frequency; the size of the set is the uncertainty layer, and the empirical coverage was verified on held-out calibration folds (Figure S13). Independent validation using mapped *Neltuma* individuals from field plots was used to assess plant-level detection performance and omission and commission errors beyond pixel-based metrics (Table S9).

To support interpretation of species co-occurrence patterns, we assessed potential *Neltuma* interactions with *V. erioloba* by denoising final drone classifications using the sieve filter in terra (v1.7-37, Hijmans 2023), converting classified patches >1 m² to polygons, and analysing nearest-neighbour distances among woody species polygons using sf (Pebesma 2018; Pebesma and Bivand 2023). We visually verified results for *V. erioloba* against photointerpretation from the RGB orthomosaics.

## 2.7 Sub-pixel *Neltuma* cover with calibrated uncertainty

Because sparse invasions rarely dominate coarse pixels, we treated continuous sub-pixel *Neltuma* cover, rather than a discrete class, as the primary landscape product. The method has four steps, each of which is validated on data held out by survey area.

*Calibrated cover target.* The averaged drone probability of *Neltuma* was recalibrated by Platt scaling on the out-of-fold predictions (pooled across sites; slope [`r values$cal_slope`]{.substituted}, intercept [`r values$cal_intercept`]{.substituted}) so that, among drone pixels assigned probability *p*, a fraction *p* is *Neltuma*. The mean of the calibrated probability over a satellite pixel footprint is then an unbiased estimate of that pixel's areal *Neltuma* cover. The calibration was applied to the drone surface before area-weighted averaging onto each satellite grid, giving one cover target per satellite pixel over every drone survey area.

*Cover regression.* Cover was regressed on the satellite predictors (spectral bands and vegetation indices) with an equal-weight ensemble of penalised regression, random forest and LightGBM; the SVM was dropped from this arm as both the least accurate and by far the slowest member. Coordinates were never predictors. The ensemble was evaluated by leave-one-site-out cross-validation, so every held-out prediction is for a survey area the model has never seen, and the out-of-fold residuals are the calibration data for the intervals below.

*Area of applicability and conformal intervals.* Each scene pixel's dissimilarity index (DI; Meyer and Pebesma 2021) is its nearest-neighbour distance to the training cells in standardised predictor space, normalised by the mean pairwise training distance and computed with a KD-tree so that it scales to the full scenes; for training cells the DI is computed against the other sites only. Per-pixel prediction intervals were obtained by DI-stratified (Mondrian) split conformal prediction: the out-of-fold residuals were binned into five DI strata, the finite-sample conformal quantile of the absolute residual was taken within each stratum, and every scene pixel received the half-width of its own stratum, so intervals widen automatically with dissimilarity. The area of applicability (AOA) was tied to the guarantee rather than to a fixed outlier rule: it extends from low DI up to the last DI band in which the intervals' out-of-fold coverage still met a floor of 85%, capped at the 99th percentile of training DI. Because leave-site-out folds break the exchangeability a finite-sample coverage theorem requires, coverage is reported as an empirical result rather than assumed: the stratum quantiles were re-estimated with each site held out in turn and evaluated on that site, and we report both the pooled and the per-site coverage (Section 3.4; Table S13).

*Landscape area.* *Neltuma* area is the sum of predicted cover over pixels within the AOA. The aggregate was bias-corrected by a prediction-powered rectifier (Angelopoulos et al. 2023) measured on the held-out drone cells and stratified by predicted cover (bands at 2, 5, 10 and 25%), so the correction applied to dense pixels is learned from dense cells rather than smeared from the sparse majority; its 95% confidence interval is a cluster bootstrap over the seven survey areas, the unit of spatial replication. The same cover surfaces were summarised to 250 m hexagons and thresholded into invasion phases (Table S8; after Müllerová et al. 2023), with each hexagon's phase also evaluated at the lower and upper 90% conformal bounds to give a range of area per phase. Spatial patterns of cover were explored descriptively in relation to distance from settlements and roads (OpenStreetMap), without formal inference.

::: {.pipeline-pending}
Figure 2 (workflow) is an authoring item. Proposed content: the two linked pipelines with the new elements marked: kNNDM evaluation with W; equal-weight probability averaging; Platt-calibrated drone probability → area-averaged cover target; cover ensemble on leave-site-out folds; DI/AOA; DI-stratified conformal intervals; prediction-powered area with site bootstrap; cover phases with ranges.
:::

Figure 2. Hierarchical machine-learning workflows integrate drone and satellite observations for multi-scale mapping of sparse *Neltuma* invasion. The drone workflow compiles predictor and training data, tunes and benchmarks learners in mlr3 under prediction-domain-matched spatial cross-validation, and produces averaged, calibrated *Neltuma* probability surfaces. The satellite workflow uses these surfaces as sub-pixel cover targets, regresses cover on satellite predictors, and attaches an area of applicability, conformal prediction intervals and a prediction-powered area estimate.

# 3. Results

## 3.1 Drone data enable accurate detection of sparse *Neltuma* at local scales in the Kalahari

Classifications across the seven drone survey areas achieved high performance, with overall accuracies of [`r values$site_best_min`]{.substituted}–[`r values$site_best_max`]{.substituted}% for the best learner per site under spatial cross-validation. Final classification maps for all survey areas are shown in Figure 4A and Figures S1–S7, with an example from Bokspits 1 in Figure 3.

::: {.pipeline-pending}
Figure 3 (orthomosaic detail and photographs) is an authoring item; original caption retained.
:::

Figure 3. Drone multispectral classification resolves sparse *Neltuma* and species associations at plant scales in heterogeneous Kalahari vegetation. (A) Example classification from the Bokspits 1 survey area showing fine-scale discrimination of *Neltuma*, *V. erioloba*, *R. trichotomum*, grass and bare sand. (B) Field photograph illustrating *Neltuma* occurring beneath a *V. erioloba* canopy on a dune margin. (C) True-colour orthomosaic detail with field training points. (D) Corresponding classified close-up showing species-level discrimination, including minor edge misclassification at canopy margins.

Benchmarking across survey areas and predictor stacks showed consistently strong performance and little to choose between learners. Mean overall accuracy across the 28 site × stack tasks ranged from [`r values$learner_bottom_pct`]{.substituted}% to [`r values$learner_top1_pct`]{.substituted}% ([`r values$learner_top1`]{.substituted}), the equal-weight probability average scored [`r values$average_pct`]{.substituted}%, and in none of the 28 tasks did the winning learner beat the runner-up by more than the winner's own fold-to-fold standard deviation (Figure 4B). Most winning SVMs chose a linear kernel and the penalised multinomial matched them, indicating that class boundaries in this feature space are close to linear and that predictor composition, not model complexity, carries the signal. Predictor stacks were likewise close: mean accuracy of the per-site winners rose from [`r values$stack_min_pct`]{.substituted}% with the five spectral bands alone to [`r values$stack_best_pct`]{.substituted}% with the CHM and all vegetation indices ([`r values$stack_best`]{.substituted}), and *Neltuma* recall was highest ([`r values$neltuma_drone_pct`]{.substituted}%) with multispectral bands and the CHM. On the stack used for mapping, the averaged model's own held-out accuracy was [`r values$avg_acc_min`]{.substituted}–[`r values$avg_acc_max`]{.substituted}% across sites with *Neltuma* recall of [`r values$avg_neltuma_recall_min`]{.substituted}–[`r values$avg_neltuma_recall_max`]{.substituted}%. Principal component analysis and band-level boxplots further indicated strong spectral separability between *Neltuma* and other vegetation classes (Figures S8–S9). At 90% nominal coverage the conformal prediction sets averaged [`r values$drone_setsize_min`]{.substituted}–[`r values$drone_setsize_max`]{.substituted} classes per pixel, so the drone model is close to decisive at plant scale (Figure S13).

Most *Neltuma* plants were correctly classified at pixel level, although shaded canopy edges were occasionally confused with *R. trichotomum* and *V. erioloba* (Figure 3D). Independent plant-scale validation showed 84.7% detection accuracy (Table S9), with reliable detection across a range of plant heights and growth stages. Phenological variation, from leaf-off to flowering states, was the principal source of error. Misclassification occurred in 8.7% of cases, typically involving shaded or senescent plants, while 3.8% of *Neltuma* individuals were misclassified as *V. erioloba*, all in flower.

::: {.pipeline-pending}
The plant-scale validation figures (84.7%, n = 184, 8.7%, 3.8%) are the submitted manuscript's, from the two-plot survey, which is not a pipeline input. The point layer in the archive (All_points_buffered_additional) is the classifier's own field polygons, so the pipeline's Table S9 is a resubstitution check, not an independent validation, and is reported as such in the Supplement. The independent plot survey should be added to the inputs so this number is regenerated against the averaged surfaces.
:::

Beyond classification performance, *Neltuma* occurred beneath or within 1 m of the canopy of 63% of *V. erioloba* trees (n = 192), indicating frequent co-occurrence consistent with native tree canopies acting as loci of invasion, with implications for interpreting invasion dynamics and targeting management.

```{r fig4-panels, out.width="100%"}
knitr::include_graphics(c(fig("fig_maps"), fig("fig_acc")))
```

Figure 4. Drone multispectral classification achieved ~[`r values$stack_best_pct`]{.substituted}% overall accuracy for vegetation mapping and reliable detection of sparse *Neltuma*. (A) Averaged classification surfaces for the seven survey areas. (B) Overall accuracy by predictor stack (per-site winners and their mean) and by learner across the 28 site × stack tasks, under [`r values$n_outer_iterations`]{.substituted} iterations of prediction-domain-matched spatial cross-validation; the equal-weight probability average is shown beside the individual learners.

## 3.2 Discrete satellite classification is an unreliable route to landscape extent

Cross-scale calibration of a discrete classification confronts a basic constraint: *Neltuma*'s low fractional cover within coarse pixels. Across the seven drone survey areas only [`r values$fig5_s2_pure_pct`]{.substituted}% of Sentinel-2 pixels reached the 65% purity threshold used to label them, and the share of pixels with any *Neltuma* at all rose from [`r values$fig5_wv2_any_pct`]{.substituted}% at 1.6 m to [`r values$fig5_s2_any_pct`]{.substituted}% at 10 m while the share dominated by *Neltuma* fell (Figure 5). As pixel size increases, the sparse invasion becomes a sub-pixel phenomenon almost everywhere: a classifier trained on pure pixels sees a vanishing and unrepresentative fraction of the cross-scale information, and a mixed pixel has no correct single class.

```{r fig5-panels, out.width="70%"}
knitr::include_graphics(fig("fig_cover"))
```

Figure 5. Coarser sensor grain sharply reduces the availability of high-*Neltuma*-cover pixels for training and detection. Distribution of sub-pixel *Neltuma* cover within (A) 1.6 m WorldView-2, (B) 3 m PlanetScope and (C) 10 m Sentinel-2 pixels over the drone survey areas; the dashed line is each sensor's purity threshold for discrete-class labels, and the annotation gives the share of pixels containing any *Neltuma* and the share reaching the threshold.

Evaluated as a prediction of unsurveyed parts of the landscape, the discrete satellite classifications were substantially less accurate than conventional evaluation suggests, and their ranking depended on how they were evaluated. With prediction-domain-matched folds the best WorldView-2 learner reached [`r values$hc_wv2_purity_pct`]{.substituted}% overall accuracy when trained on drone-derived labels and [`r values$hc_wv2_field_pct`]{.substituted}% on field observations alone, so incorporating drone-derived calibration data did not raise the accuracy of the discrete classification; PlanetScope and Sentinel-2 reached [`r values$hc_planet_purity_pct`]{.substituted}% and [`r values$hc_s2_purity_pct`]{.substituted}%, though on fewer classes and, for Sentinel-2, only [`r values$hc_s2_ntrain`]{.substituted} training pixels (Table S11). The same WorldView-2 task evaluated with random cross-validation returned about 75%, and with coordinate-clustered folds the accuracy rose monotonically with the number of folds, from about 49% with 5 folds to about 71% with 40 (Table S11), because many small folds hold out patches within surveyed areas whereas few large folds hold out whole areas. A single accuracy figure for a satellite classification of this kind is therefore not interpretable without its fold geometry, and the honest figure for the question the map answers is the lower one. Conformal prediction sets make the same point in map form: to keep 90% coverage the WorldView-2 classification had to offer [`r values$hc_wv2_setsize90`]{.substituted} of its six classes per pixel on average, against [`r values$drone_setsize_max`]{.substituted} or fewer for the drone (Figure S13).

Independent comparison with the drone classifications inside the survey areas revealed errors that internal validation cannot show (Table S10). The WorldView-2 classification recovered [`r values$px_wv2_recall_pct`]{.substituted}% of drone-mapped *Neltuma* pixels at [`r values$px_wv2_precision_pct`]{.substituted}% precision and matched the drone *Neltuma* extent to within [`r values$px_wv2_ratio_pct`]{.substituted}%, with [`r values$px_wv2_woody_pct`]{.substituted}% of drone-mapped *Neltuma* pixels labelled as *V. erioloba* or other woody vegetation. PlanetScope and Sentinel-2 recovered [`r values$px_planet_recall_pct`]{.substituted}% and [`r values$px_s2_recall_pct`]{.substituted}% of drone *Neltuma* pixels but at only [`r values$px_planet_precision_pct`]{.substituted}% and [`r values$px_s2_precision_pct`]{.substituted}% precision, over-predicting *Neltuma* extent by factors of [`r values$px_planet_ratio`]{.substituted} and [`r values$px_s2_ratio`]{.substituted} (Figure 6). The single-model landscape totals were correspondingly unstable: across the five tuned learners the Sentinel-2 scene's *Neltuma* area ranged from [`r values$hc_s2_area_min`]{.substituted} to [`r values$hc_s2_area_max`]{.substituted} ha (Table S14). A discrete classification of a sparse target at coarse grain thus yields a landscape extent that is neither accurate nor stable, which motivated the sub-pixel cover product; the discrete WorldView-2 landscape map and the invasion phases derived from it are retained for comparison in Figure S14.

## 3.3 Detection of sparse *Neltuma* in the Kalahari is constrained by satellite image grain

Increasing sensor grain reduced *Neltuma* detectability, increased confusion with other classes and caused progressive loss of spatial detail in mapped invasion patterns (Figure 6). Against the drone reference, precision of the *Neltuma* class fell from [`r values$px_wv2_precision_pct`]{.substituted}% at 1.6 m to [`r values$px_planet_precision_pct`]{.substituted}% at 3 m and [`r values$px_s2_precision_pct`]{.substituted}% at 10 m, and the ratio of satellite to drone *Neltuma* extent rose from [`r values$px_wv2_ratio`]{.substituted} to [`r values$px_planet_ratio`]{.substituted} and [`r values$px_s2_ratio`]{.substituted} (Figure 6E). Overall agreement with the drone maps did not fall in step ([`r values$px_wv2_agree_pct`]{.substituted}%, [`r values$px_planet_agree_pct`]{.substituted}% and [`r values$px_s2_agree_pct`]{.substituted}%), because the coarser sensors classify fewer classes and the abundant bare-ground and grass classes dominate agreement. This divergence shows that conventional overall accuracy can mask substantial losses in sensitivity to sparse ecological targets, and that at 3–10 m a discrete map of *Neltuma* mainly maps mixed pixels that contain some *Neltuma*.

```{r fig6-panels, out.width="100%"}
knitr::include_graphics(fig("fig_sensors"))
```

Figure 6. Detectability of sparse *Neltuma* declines sharply as sensor grain becomes coarser. Discrete classifications of the same area from (A) drone multispectral imagery (7 cm), (B) WorldView-2 (1.6 m), (C) PlanetScope (3 m) and (D) Sentinel-2 (10 m), illustrating progressive loss of spatial detail and increasing confusion among vegetation classes with coarser resolution. (E) Pixel-level recall and precision of the *Neltuma* class against the drone classifications within the survey areas, and the mean conformal prediction-set size at 90% coverage, per sensor.

## 3.4 Sensor grain determines where sub-pixel *Neltuma* cover can be mapped with confidence

Modelling *Neltuma* as continuous sub-pixel cover, rather than a discrete class, mapped a widespread but overwhelmingly low-cover invasion and, more importantly, revealed how sensor grain controls not only detection but the extent over which detection can be trusted (Figure 7). Within its area of applicability the WorldView-2 model estimated [`r values$cover_wv2_ha`]{.substituted} ha of *Neltuma* cover (95% CI [`r values$cover_wv2_lo`]{.substituted}–[`r values$cover_wv2_hi`]{.substituted} ha) over [`r values$cover_wv2_aoa_pct`]{.substituted}% of the study area, with PlanetScope [`r values$cover_planet_ha`]{.substituted} ha ([`r values$cover_planet_lo`]{.substituted}–[`r values$cover_planet_hi`]{.substituted}) over [`r values$cover_planet_aoa_pct`]{.substituted}% and Sentinel-2 [`r values$cover_s2_ha`]{.substituted} ha ([`r values$cover_s2_lo`]{.substituted}–[`r values$cover_s2_hi`]{.substituted}) over [`r values$cover_s2_aoa_pct`]{.substituted}% (Figure 8A; Table 2). The three estimates are of similar magnitude with overlapping confidence intervals, in contrast to the discrete classifications, whose extents differed several-fold. Mean predicted cover within the area of applicability was [`r values$cover_wv2_mean_pct`]{.substituted}% for WorldView-2, and the aggregate bias measured on the held-out drone cells was small ([`r values$cover_wv2_bias_pp`]{.substituted} percentage points).

Held-out performance at pixel level was modest and grain-dependent, as expected for a target whose cover averages a few percent: leave-site-out RMSE of cover was [`r values$cover_wv2_rmse_pp`]{.substituted}, [`r values$cover_planet_rmse_pp`]{.substituted} and [`r values$cover_s2_rmse_pp`]{.substituted} percentage points for WorldView-2, PlanetScope and Sentinel-2, with the coarsest grain the smoothest and least noisy per pixel and the finest grain the most spatially detailed (Table S13). The DI-stratified conformal intervals earned their nominal level in aggregate: with each site's intervals calibrated on the other six sites, pooled 90% coverage was [`r values$cover_wv2_cov90_honest`]{.substituted}%, [`r values$cover_planet_cov90_honest`]{.substituted}% and [`r values$cover_s2_cov90_honest`]{.substituted}% (Figure 8B), with a mean 90% interval width of [`r values$cover_wv2_width90_pp`]{.substituted} percentage points for WorldView-2. Coverage was not uniform across sites, however, ranging from [`r values$cover_wv2_cov90_site_min`]{.substituted}% to [`r values$cover_wv2_cov90_site_max`]{.substituted}% for WorldView-2. The shortfall is concentrated at the one densely invaded survey area (Struizendam 4, true cover [`r values$dense_site_truth_pct`]{.substituted}%), which every sensor under-predicted by [`r values$cover_wv2_dense_bias_pp`]{.substituted}–[`r values$cover_planet_dense_bias_pp`]{.substituted} percentage points when it was held out (Table S13): a model trained without any dense invasion cannot extrapolate to it, and its intervals do not fully cover it. This is the single largest source of uncertainty in the landscape totals and is why their confidence intervals are wide and asymmetric.

The mapped cover is not, therefore, mapped with equal confidence everywhere, and it is grain that decides where. The densest *Neltuma* tracks the ephemeral river valley and the settlement and road corridors around Struizendam, Vaalhoek and Bokspits (Figure 7A). At WorldView-2's 1.6 m grain these corridors fall within the area of applicability and are mapped with the coverage guarantee intact (Figure 7B, green), but at coarser grains they are increasingly flagged as beyond applicability (Figure 7B, purple): mixed 3–10 m pixels along the corridors are spectrally dissimilar from the sparse-matrix-dominated training, so the coarse-sensor models are extrapolating precisely where the invasion is most severe. This reflects grain rather than sampling effort: from the same drone surveys each grain yields a very different training density ([`r values$cover_wv2_ntrain`]{.substituted}, [`r values$cover_planet_ntrain`]{.substituted} and [`r values$cover_s2_ntrain`]{.substituted} labelled cells), and only where the grain resolves individual canopies do dense-corridor pixels resemble the dense training cells. Mapping the dense invasion fronts at coarse grain would therefore require training that densely spans the invasion gradient, whereas at fine grain the resolution does that work.

The two uncertainty views also distinguish the coarse sensors. Although Sentinel-2's applicability is the narrowest (Figure 7B), its conformal interval widths beyond the area of applicability are no larger, and in places smaller, than PlanetScope's (Figure 7C–D), whose scene-composite radiometry imprints spatially patterned error across the interior of the study area and partly inflates its recovered area. Grain and radiometric consistency, not resolution alone, govern where a coarse sensor's predictions can be trusted.

```{r fig7-grain, out.width="100%"}
knitr::include_graphics(fig("fig_cover_grain"))
```

Figure 7. Sensor grain governs both *Neltuma* cover and its trustworthy extent. Columns: WorldView-2 (1.6 m), PlanetScope (3 m), Sentinel-2 (10 m); roads and the three villages overlaid. (A) Sub-pixel *Neltuma* cover within each sensor's coverage-driven area of applicability (grey = beyond applicability; the un-masked cover surface is Figure S15). (B) Dissimilarity index relative to the AOA threshold (green within, purple beyond): the dense corridors stay within applicability at fine grain but are pushed beyond it at coarse grain. (C) Conformal 90% prediction-interval width (the pixel-level cover uncertainty). (D) Detection confidence as a bivariate of cover (red) against DI-driven extrapolation uncertainty (blue): confident *Neltuma* appears red.

```{r fig8-panels, out.width="100%"}
knitr::include_graphics(c(fig("fig_cover_area"), fig("fig_cover_coverage")))
```

Figure 8. Sub-pixel cover area and interval calibration. (A) Bias-corrected *Neltuma* cover area per sensor (point = prediction-powered estimate; bar = 95% site-cluster-bootstrap CI; cross = naive sum of fractions). (B) Empirical coverage of the DI-stratified conformal intervals against the nominal level, with each site's intervals calibrated on the other six sites: pooled coverage per sensor (lines) and per-site coverage (points); the diagonal indicates calibrated intervals.

Table 2. Sensor evaluation summary. Grain, labelled training cells for the cover model, mean Wasserstein *W* of the satellite evaluation folds, pixel-level *Neltuma* recall and precision of the discrete classification against the drone maps, mean conformal set size at 90% coverage, leave-site-out RMSE of sub-pixel cover, pooled 90% interval coverage, area of applicability as a share of the study area, and bias-corrected *Neltuma* cover area with its 95% CI.

```{r table2-sensors}
knitr::kable(sensor_tab, row.names = FALSE, digits = c(0, 1, 0, 1, 0, 0, 2, 1, 1, 0, 0, 0, 0),
             col.names = c("Sensor", "Grain (m)", "Cover training cells", "W (km)",
                           "Neltuma recall (%)", "Neltuma precision (%)", "Set size @90%",
                           "Cover RMSE (pp)", "Coverage @90% (%)", "AOA (%)",
                           "Area (ha)", "CI low", "CI high"))
```

## 3.5 *Neltuma* invasion is widespread and sparse, and concentrates along settlements, roads and river corridors

Across the mapped area the invasion is dominated by low-density fronts rather than dense stands. Partitioning the WorldView-2 cover surface into invasion phases placed [`r values$cover_wv2_expansion_pct`]{.substituted}% of the study area in active expansion (1.5–15% cover) and [`r values$cover_wv2_incursion_pct`]{.substituted}% in initial incursion, with only [`r values$cover_wv2_dominance_pct`]{.substituted}% reaching dense dominance (Table 3). Because the phase thresholds sit close to the model's per-pixel uncertainty, the phase of an individual hexagon is not certain, and the ranges in Table 3 show how the split moves when every hexagon is evaluated at the lower and upper bounds of its 90% conformal interval: the extent of dominance is bounded above by [`r values$cover_wv2_dominance_hi_pct`]{.substituted}% under the upper bound, while under the lower bound [`r values$cover_wv2_preincursion_lo_pct`]{.substituted}% of the landscape cannot be distinguished from pre-incursion. The robust statements are therefore that dense invasion is confined to a small fraction of the landscape and that sparse invasion is very widespread.

Both invasion density and maturity increased in and around settlements and along roads and declined with distance from these features: mean WorldView-2 cover of hexagons within the area of applicability was [`r values$cover_wv2_sett_near_pct`]{.substituted}% within 1 km of a village against [`r values$cover_wv2_sett_far_pct`]{.substituted}% at 3–6 km, and [`r values$cover_wv2_road_near_pct`]{.substituted}% within 250 m of a road against [`r values$cover_wv2_road_far_pct`]{.substituted}% at 1–3 km (Figures S10–S11). These are descriptive patterns rather than tested relationships, and roads, settlements and livestock infrastructure are concentrated along the dry riverbeds, so the features are strongly collinear; the pattern is consistent with field observations implicating livestock as a key vector of *Neltuma* spread.

Table 3. The mapped *Neltuma* invasion is overwhelmingly in expansion and early-incursion phases, with negligible dense dominance (WorldView-2 sub-pixel cover in 250 m hexagons; Table S8 thresholds). Point = phase of the predicted cover; low and high = the phase split when every hexagon is evaluated at the lower and upper bounds of its 90% conformal interval.

```{r table3-phases}
p3 <- phase_tab[phase_tab$sensor == "wv2", c("phase", "pct_of_area", "pct_lower", "pct_upper", "area_ha")]
knitr::kable(p3, row.names = FALSE, digits = c(0, 1, 1, 1, 0),
             col.names = c("Invasion phase", "% of area (point)", "% at lower bound", "% at upper bound", "Area (ha, point)"))
```

# 4. Discussion

Exotic invasive plants such as *Neltuma* threaten African rangelands by degrading ecosystem services for people and wildlife, yet detecting sparse invasions remains challenging because invasion patterns often occur below the grain of operational Earth observation systems. This study shows that sparse invasion detection is fundamentally a scale-mismatch problem, but that integrating field observations, drone imagery and satellite data within a hierarchical multi-scale framework can turn the mismatch into a training signal, quantify sensor-grain constraints and support management-relevant mapping of invasion dynamics with stated uncertainty. In the Kalahari, this approach revealed both the value of fine-resolution ecological information for strengthening landscape-scale inference and the limitations of coarser satellite systems for detecting low-cover invasions at expanding invasion fronts, and it showed that where a map can be trusted is as much a property of the sensor as what it detects.

## 4.1 Multispectral drone observations provide accurate plant-scale detection of sparse *Neltuma*

Multispectral drone observations provided accurate plant-scale detection of sparse *Neltuma*, demonstrating the capacity of fine-resolution Earth observation to resolve invasion signals that are often difficult to detect with coarser sensors (Slade et al. 2023). High classification performance across diverse growth forms indicates that combining spectral and structural information can discriminate *Neltuma* effectively within heterogeneous dryland vegetation, although detection was weaker for leaf-off individuals, highlighting phenology as a persistent source of uncertainty. Improved performance when vegetation indices were included reflects the value of phenologically sensitive predictors in dryland systems, while canopy height information reduced confusion with *V. erioloba* and other woody species by capturing structural differences not resolved by spectral information alone. These results were obtained using a pixel-based classification framework and therefore include some misclassification associated with canopy margins, mixed pixels and shadowed areas that may be reduced by object-based approaches incorporating spatial context.

Although all algorithms performed well, no learner won by more than the fold-to-fold noise, and a penalised multinomial regression matched the SVM and tree ensembles, indicating near-linear class boundaries in this feature space and that predictor composition and robust evaluation were more important than classifier choice. This finding has a practical consequence that we adopted: because equally accurate models draw materially different maps of a class that occupies about 1% of the landscape, we map the equal-weight average of the tuned learners' probabilities rather than a nominal winner, which reduced the variance of sparse-class areas at no cost in accuracy. The prediction-domain-matched spatial cross-validation framework was particularly important because spatially structured observations lead to overly optimistic estimates of predictive performance when conventional random train–test splits are used (Roberts et al. 2017), while over-blocking a small site penalises a model for extrapolation it will never be asked to do (Linnenbrink et al. 2024). Building folds to match the prediction situation, and reporting the mismatch statistic *W*, resolves that tension explicitly and provides a more realistic assessment of how well classifications generalise to the locations they map (Milà et al. 2022). Considering variation across repeated fold constructions, rather than mean accuracy alone, provided a more informative assessment of model stability across spatial contexts, and the conformal prediction sets show that the drone model is close to decisive at plant scale.

These findings demonstrate that drone observations are a powerful ecological observation layer for detecting sparse invasions and characterising invasion structure at local scales. While object-based approaches may further improve crown delineation and reduce local pixel-level errors, the pixel-based framework adopted here was sufficient to achieve accurate plant-scale detection and to support the subsequent cross-scale analyses. However, their principal value in this study lies not as a stand-alone monitoring solution, but as the fine-resolution foundation of a hierarchical multi-scale framework. Given the spatial extent of invasion in landscapes such as the 105,200 km² Kgalagadi District, drone surveys alone do not provide operationally scalable monitoring, but they can supply the ecological detail and calibration needed to strengthen broader-scale satellite mapping.

## 4.2 Drone data calibrate satellite estimates of sub-pixel *Neltuma* cover, and honest evaluation changes what can be claimed

The contribution of drone data to satellite mapping proved to be different in kind from the one usually claimed. Used as a source of purer training labels for a discrete classification, drone-derived calibration did not improve the WorldView-2 classification once accuracy was measured as the ability to predict unsurveyed parts of the landscape; and that classification's accuracy itself moved by more than 20 percentage points depending on how the evaluation folds were drawn, with random cross-validation returning the friendliest figure. Independent comparison with the drone maps then showed that the discrete satellite classifications, particularly at 3–10 m, over-predicted *Neltuma* extent several-fold with low precision, and that their landscape totals varied by tens of percent between equally accurate learners. These results caution against reporting a single cross-validated accuracy for a satellite classification of a sparse target, and against deriving area from a discrete map of a sub-pixel phenomenon. They also make concrete the costs of both error types for management: a map that over-predicts extent several-fold would direct removal effort to mixed pixels that contain little *Neltuma*, while a map that recovers only three fifths of the drone-mapped plants at 1.6 m would leave sparse incursions untreated.

Motivated by these limitations, we moved beyond assigning each coarse pixel to a single discrete class and instead modelled *Neltuma* as continuous sub-pixel cover (Frazier and Wang 2011; Shiferaw et al. 2019), calibrating drone-derived probabilities to fractional cover and regressing this target onto the satellite predictors. Here the drone surveys contribute what a field campaign cannot: a complete, area-weighted cover label for every satellite pixel they overlap, hundreds of thousands of them at fine grain, rather than a few hundred pure pixels. This framing better captures the spatial heterogeneity of mixed pixels, where confidence in species presence varies continuously, and yielded concordant landscape estimates across all three sensor grains, in contrast to the discrete classification. It contributes to a small but growing body of work integrating observations across spatial scales, with geolocated fine-scale data from UAVs used to improve classification and prediction in agricultural applications (Maimaitijiang et al. 2020; Ogungbuyi et al. 2024), strengthen aboveground biomass estimation (Cheng et al. 2024; Kolluru et al. 2026), and support invasive plant mapping (Holden, Rebelo, and New 2021; Page et al. 2022), although comparable hierarchical calibration frameworks remain rare in sparse invasive plant detection (Singh, Surasinghe, and Frazier 2024).

To make the resulting maps decision-ready, we paired the cover regression with distribution-free conformal prediction (Angelopoulos and Bates 2022): we delimited an area of applicability from a dissimilarity index, attached per-pixel prediction intervals stratified by that index, validated their coverage on held-out survey areas, and reported aggregate invasion area with a cover-stratified, prediction-powered bias correction (Angelopoulos et al. 2023) and an explicit confidence interval. Two honest limitations of that validation deserve emphasis. Pooled coverage met the nominal level, but coverage at the one densely invaded survey area did not, because no other site contains comparable density and the model under-predicts a regime it has not seen; and the leave-site-out design, with seven survey areas, gives the landscape totals wide confidence intervals dominated by between-site variation. Both are properties of the sampling design rather than of the method, and both are exactly the information a manager needs: the totals are sound in aggregate, the dense stands are the least certain part of the map, and the next drone survey would reduce that uncertainty most if it sampled dense invasion. Together these steps convert a single-class extent map into a continuous, uncertainty-aware estimate of invasion extent better suited to sparse targets, providing a template for reporting not only where invasion is mapped but how confident that mapping is.

Framing detection through an explicit area of applicability also reframes the choice of sensor and where to invest survey effort. The dissimilarity index rises steeply for the mixed pixels of dense invasion fronts, so the trustworthy domain of a coarse sensor systematically excludes exactly the areas of greatest management concern, the settlement and river corridors, unless the training densely spans the invasion gradient there. This is a property of grain rather than of survey effort alone: from identical drone surveys WorldView-2 yielded [`r values$cover_wv2_ntrain`]{.substituted} labelled cells against only [`r values$cover_s2_ntrain`]{.substituted} for Sentinel-2, and only the finer grain resolved individual canopies well enough to keep the dense corridors within applicability. The applicability map is therefore not merely a caveat but an actionable one: its beyond-applicability zones identify where additional fine-resolution survey would most extend reliable coarse-sensor mapping, giving a concrete, model-driven criterion for targeting UAV acquisition rather than surveying at random. Notably, the coarse sensors do not fail in the same way. Sentinel-2's area of applicability is narrower than PlanetScope's, yet its prediction intervals beyond that domain are no wider, whereas PlanetScope's scene-composite radiometry imprints spatially structured error across the interior; sensor grain and radiometric consistency thus interact, and a slightly coarser but radiometrically stable sensor may extrapolate more gracefully than a finer but multi-scene one. Together these observations argue for reporting invasion maps alongside their applicability and interval widths, and for treating those uncertainty layers as inputs to survey design rather than as afterthoughts.

More broadly, these findings illustrate how hierarchical multi-scale workflows can bridge scale mismatches in Earth observation. Drone surveys alone are unlikely to support monitoring across landscapes of this extent, while coarser satellite systems may lack sufficient resolution to reliably detect sparse invasion fronts. A targeted UAV calibration layer can link these scales, allowing detailed observations to inform spatially extensive satellite mapping without requiring fine-resolution surveys across the entire landscape. Rather than treating UAV acquisition as a one-off calibration exercise, future recursive approaches could iteratively target surveys towards locations where additional fine-resolution observations are expected to provide the greatest reduction in model uncertainty, which the area-of-applicability and interval-width layers now make explicit. This could provide a practical optimisation strategy for allocating UAV effort, concentrating surveys in representative locations, areas of particular ecological interest, or regions where model uncertainty is greatest, while reducing the spatial coverage and logistical requirements of systematic drone mapping across the full monitoring area (Holden, Rebelo, and New 2021). Such targeted acquisition may offer a practical compromise between the ecological detail of fine-resolution observations and the spatial coverage required for operational monitoring.

## 4.3 Detection of sparse *Neltuma* depends on matching sensor grain to canopy structure

Our results show that accurate detection of sparse *Neltuma* depends fundamentally on the correspondence between sensor grain and the spatial structure of the invasion target. Fine-resolution WorldView-2 imagery (1.6 m) enabled useful and operationally relevant mapping of *Neltuma* at scales suitable for district- and protected-area management, whereas at 3 m and 10 m the discrete classification lost precision sharply as pixel size increasingly exceeded the characteristic scale of sparse canopies and low-density invasion patches. This supports the interpretation that reduced performance at coarser grains reflects a scale-mismatch problem, in which mixed pixels obscure invasion signals and erode sensitivity to low-cover occurrences at invasion fronts.

The steep decline in *Neltuma* precision relative to stable overall agreement further emphasises that conventional accuracy metrics can mask losses in sensitivity to sparse ecological targets. Modelling cover rather than class changes the character of the constraint without removing it: all three grains recover a similar landscape total of cover, but the coarser grains do so with a narrower area of applicability that excludes the densest invasion, so the grain limit re-appears as a limit on where, rather than whether, the coarse sensor can be trusted. These findings reinforce the importance of matching observation grain to organismal pattern when using Earth observation for invasive plant detection and extend broader evidence that spatial scale is fundamental to understanding and managing biological invasions (Slade et al. 2023; McMillan et al. 2023; Müllerová et al. 2023). They are also consistent with recent synthesis highlighting spatial resolution as a critical limitation for detecting young trees and sparse stands of *Neltuma* at invasion fronts (Eckert et al. 2025).

An important practical implication is that optical data at 10 m resolution or coarser, including systems such as Sentinel-2 and especially historical archives such as Landsat, are likely to yield highly uncertain detection of sparse and discontinuous invasions, and should be used with an explicit area of applicability and interval rather than as a discrete map. This constrains not only current mapping but also retrospective analyses seeking to reconstruct invasion dynamics from long-term satellite records (Mbaabu et al. 2019). Where fine-resolution imagery is unavailable, improving inference will therefore require approaches that explicitly account for the scale-dependent uncertainty and loss of spatial detail inherent in coarser observations; the sub-pixel cover product with its uncertainty layers is one such approach. More broadly, these results suggest that managing scale mismatch is not simply a technical challenge but a central ecological requirement for reliable invasion mapping (Vaz et al. 2018). Where fine-resolution commercial imagery may be prohibitively costly at national scales, the challenge is therefore not only to acquire finer data, but to develop cost-effective and scalable modelling and monitoring approaches that retain useful information about sparse phenomena at coarser grains. Addressing this remains a key frontier for extending Earth observation approaches to early detection and monitoring of dryland invasions.

## 4.4 *Neltuma* is expanding across the study area, with broad potential for further densification

The spatial pattern of *Neltuma* distribution is consistent with livestock-mediated dispersal as an important invasion pathway in the western Kalahari, in line with patterns observed in other semi-arid rangelands (Campos et al. 2011; Kebede and Coppock 2015; Shiferaw et al. 2019). Elevated cover around settlements, roads and dune valleys aligns with livestock movement patterns, whereby animals concentrate near villages before ranging widely into surrounding savanna (Mosweu et al. 2013; Kamiri, Choge, and Becker 2024; Eschen et al. 2024; Meroni et al. 2017), although roads, settlements and livestock infrastructure are concentrated along the ephemeral riverbeds and the pattern alone cannot separate these covariates. Since its deliberate introduction in 1980 (van der Poel and Timberlake 1980), *Neltuma* appears to have expanded through coupled processes of local densification and outward dispersal. Although much of the study area remains characterised by low-density infestations, this trajectory suggests substantial potential for continued expansion and densification without effective intervention.

Interactions between livestock sector expansion and episodic drought may further accelerate spread. *Neltuma* is already present in many villages across the western Kalahari (Figure S12), and patterns observed here may foreshadow broader regional dynamics as cattle posts, boreholes and ranches continue expanding westwards (Keeping et al. 2019). Local community accounts of pulse-like invasion during drought support a mechanism in which forage scarcity increases pod consumption and extends livestock ranging into savanna landscapes. Although cattle movement from infested areas into northern Botswana is constrained by foot-and-mouth disease controls (Kabelo et al. 2023), other domestic stock, including goats, sheep, donkeys and horses, may facilitate longer-distance dispersal. Increased livestock mobility, temporary water provisioning and policies promoting grazing expansion risk inadvertently reinforcing invasion processes (Chilume, Keotshepile, and Macinnis-Ng 2025).

Current management strategies centred on utilisation are unlikely to provide an effective or economically viable response at landscape scale (Mbaabu et al. 2019). Although Botswana's Integrated Mesquite (Prosopis Species) Management Strategy (2022) promotes control through uses such as charcoal and livestock feed, these approaches depend on sufficiently dense infestations to offset harvesting, transport and processing costs. Yet the invasion mapped here occurs overwhelmingly at low cover across large areas, with dense dominance confined to well under 1% of the landscape even at the upper bound of its uncertainty, precisely the configuration in which utilisation is least economically viable. The BORAVAST community charcoal initiative within the study area illustrates these constraints: *Neltuma* occurred predominantly as extensive, low-density infestations interspersed with relatively small high-density patches, increasing harvesting, transport and kiln deployment costs while limiting economies of scale. This illustrates how invasion pattern, as well as total invaded area, can constrain utilisation-based management in remote dryland landscapes.

More fundamentally, utilisation is unlikely to provide a stand-alone solution if the primary management objective is to prevent further spread and densification. The extensive low-density invasion mapped here represents an important window for intervention, but also the stage least suited to biomass-based utilisation. Early-stage removal across remote dune savannas will inevitably be logistically challenging, suggesting a need for approaches that combine targeted control with local knowledge and participation. Local herders may be well positioned to contribute to surveillance and early removal where appropriately supported and remunerated, for example through community-based approaches analogous to Botswana's Herding for Health programme (Heermans et al. 2021). Such approaches may be particularly valuable during early invasion stages, when infestations are widespread but remain relatively sparse and potentially more tractable. More broadly, effective implementation will require a two-way exchange between research, management and local practitioners, allowing spatial information and ecological evidence to inform action while ensuring that management experience and practical constraints feed back into adaptive strategies (Foxcroft et al. 2020).

Given the escalating costs and declining success of late-stage control (Shackleton et al. 2014), reliance on utilisation alone is unlikely to halt invasion. Instead our results support a spatially differentiated management strategy that prioritises early detection and removal of sparse incursions while targeting dense infestations with context-appropriate control measures within a broader community-based natural resource management framework. Spatially explicit information on invasion density, delivered with its uncertainty, could help prioritise these interventions: the cover map distinguishes areas where early removal remains feasible from established infestations where control or utilisation may be more appropriate (Shiferaw et al. 2019), the interval and applicability layers indicate where that distinction is reliable, and the cost of acting on a false positive or missing a false negative can be weighed against the stated confidence rather than assumed. Without sustained and well-resourced intervention, further expansion and densification are likely, with consequences for groundwater availability, livestock production, irrigated agriculture and associated ecosystems, echoing severe social and ecological impacts documented elsewhere in eastern and southern Africa (Ndhlovu, Milton, and Esler 2016; Malongweni et al. 2025; Eckert et al. 2025).

## 4.5 Transferability and recommendations

The framework transfers to other *Neltuma*-invaded regions and to other sparse, heterogeneous ecological targets because each of its components is generic: a fine-resolution classification that resolves the target, a calibrated probability that is area-averaged onto the coarser grid, a regression with prediction-domain-matched evaluation, and uncertainty layers that state where and how far the result can be trusted. Three recommendations follow from our results. First, treat drone surveys as the source of sub-pixel cover labels rather than of pure-pixel class labels; the former uses all of the cross-scale information and the latter discards nearly all of it. Second, evaluate satellite products against the question they answer, with folds that match the prediction distances and with the mismatch statistic reported, and expect the honest accuracy to be lower than a random or lightly blocked cross-validation suggests. Third, report an area of applicability and calibrated intervals with every landscape estimate, validate their coverage on held-out areas, and use the beyond-applicability zones to plan the next survey: for a sparse invasion the dense fronts will usually be the least well-sampled regime and the most valuable place to fly.

# 5. Conclusion

This study shows that integrating field observations, drone imagery and satellite data within a hierarchical multi-scale framework can substantially improve detection and mapping of sparse *Neltuma* invasions, while revealing important limits imposed by sensor grain and by how mapping products are evaluated. Fine-resolution drone data supported accurate plant-scale detection ([`r values$stack_best_pct`]{.substituted}% overall accuracy, 84.7% of independently mapped plants detected) and, used as calibrated sub-pixel cover labels, made it possible to estimate landscape *Neltuma* cover from satellite imagery with a validated uncertainty: about [`r values$cover_wv2_ha`]{.substituted} ha ([`r values$cover_wv2_lo`]{.substituted}–[`r values$cover_wv2_hi`]{.substituted} ha) across the 445 km² study area, overwhelmingly at low cover. A discrete satellite classification was not a reliable route to that number: under honest spatial evaluation its accuracy fell well below conventional estimates, and against the drone maps it recovered [`r values$px_wv2_recall_pct`]{.substituted}% of *Neltuma* pixels at 1.6 m and over-predicted extent by factors of [`r values$px_planet_ratio`]{.substituted} and [`r values$px_s2_ratio`]{.substituted} at 3 and 10 m. Sensor grain governed not only detection but the extent over which detection could be trusted, with the dense invasion fronts falling outside the area of applicability of the coarser sensors.

For *Neltuma* in the Kalahari, the results indicate substantial potential for continued expansion and reinforce the urgency of proactive, spatially differentiated management while infestations remain relatively sparse. More broadly, our findings demonstrate that targeted integration of fine- and coarse-resolution observations can provide more robust evidence for operational decision-making than either scale alone, provided that the evidence is delivered with its applicability and its uncertainty. For those mapping sparse invasions and other heterogeneous ecological targets, we recommend incorporating fine-resolution UAV observations as sub-pixel cover labels where they can materially improve landscape-scale inference, evaluating satellite products with prediction-domain-matched validation, and reporting calibrated intervals and an area of applicability so that maps can be used, and improved, with a known level of confidence.

# Acknowledgements

This work was supported by the Oppenheimer Programme in African Landscape Systems (OPALS), jointly funded by the University of Exeter, Sarah Turvill, and Oppenheimer Generations Research and Conservation. We thank the Government of Botswana Ministry of Environment, Natural Resources Conservation and Tourism for permission to undertake this research (permit ENT 8/36/4LII(09)). We thank the Civil Aviation Authority of Botswana for authorising the drone surveys. We thank the Dikgosi, Village Development Committees, and the communities of Bokspits, Struizendam and Vaalhoek for their generous support, hospitality and collaboration in the field research. We acknowledge the European Space Agency (ESA) for providing WorldView-2 data through its Third Party Missions programme, and the Norway International Climate and Forest Initiative (NICFI) for providing access to PlanetScope SuperDove imagery. We thank two anonymous reviewers for their constructive and thoughtful feedback, which helped us strengthen the manuscript. For the purpose of open access, the author has applied a Creative Commons Attribution (CC BY) licence to any Author Accepted Manuscript version arising from this submission.

# Open Research

The full analysis is a reproducible targets pipeline in R; the codebase, configuration and the ground-observed data for training and validation are available at https://doi.org/10.5281/zenodo.16681147 under a GPL-3.0 licence, together with a lock file of the computational environment. The processed drone data (georeferenced orthomosaics of spectral reflectance, vegetation indices, and canopy height models) for each of the seven survey areas, the drone classification and probability surfaces, and the satellite sub-pixel cover, dissimilarity-index and interval rasters are available at https://doi.org/10.5281/zenodo.18506271.

# Disclosure

Our study brings together authors from two countries, including scientists based in the country where the study was carried out. All authors were engaged with the research and study design to ensure that the diverse sets of perspectives they represent was considered. Where relevant, literature published by scientists from the region was cited. Portions of the draft manuscript text were edited with assistance from large language models (ChatGPT, OpenAI; Claude, Anthropic). All outputs were critically reviewed and revised by the authors, who take full responsibility for the content.

# References

::: {.pipeline-note}
Reference list carried over from the submitted manuscript. New citations to add: Angelopoulos et al. (2023) prediction-powered inference, *Science* 382:669–674; Meyer & Pebesma (2021) area of applicability, *Methods in Ecology and Evolution* 12:1620–1633; Meyer et al. (2024) CAST R package; Linnenbrink et al. (2024) kNNDM, *Geoscientific Model Development* 17:5897–5912; Frazier & Wang (2011) sub-pixel cover, *Remote Sensing of Environment*; Shiferaw et al. (2019) already cited. Already cited in the submitted version: Angelopoulos & Bates (2022), Milà et al. (2022), Meyer & Pebesma (2022), Roberts et al. (2017).
:::

Adam, Elhadi, Nyasha Mureriwa, and Solomon Newete. 2017. "Mapping Prosopis Glandulosa (Mesquite) in the Semi-Arid Environment of South Africa Using High-Resolution WorldView-2 Imagery and Machine Learning Classifiers." Journal of Arid Environments 145 (October): 43–51. https://doi.org/10.1016/j.jaridenv.2017.05.001.

Adar, Shay, Marcelo Sternberg, Tarin Paz-Kagan, Zalmen Henkin, Guy Dovrat, Eli Zaady, and Eli Argaman. 2022. "Estimation of Aboveground Biomass Production Using an Unmanned Aerial Vehicle (UAV) and VENμS Satellite Imagery in Mediterranean and Semiarid Rangelands." Remote Sensing Applications: Society and Environment 26 (April): 100753. https://doi.org/10.1016/j.rsase.2022.100753.

Alvarez-Taboada, Flor, Claudio Paredes, and Julia Julián-Pelaz. 2017. "Mapping of the Invasive Species Hakea Sericea Using Unmanned Aerial Vehicle (UAV) and WorldView-2 Imagery and an Object- Oriented Approach." Remote Sensing 9 (9): 9. https://doi.org/10.3390/rs9090913.

Angelopoulos, Anastasios N., and Stephen Bates. 2022. "A Gentle Introduction to Conformal Prediction and Distribution-Free Uncertainty Quantification." arXiv:2107.07511. https://doi.org/10.48550/arXiv.2107.07511.

Angelopoulos, Anastasios N., Stephen Bates, Clara Fannjiang, Michael I. Jordan, and Tijana Zrnic. 2023. "Prediction-Powered Inference." Science 382 (6671): 669–674. https://doi.org/10.1126/science.adi6000.

Assmann, Jakob J., Jeffrey T. Kerby, Andrew Cunliffe, and Isla H. Myers-Smith. 2018. "Vegetation Monitoring Using Multispectral Sensors — Best Practices and Lessons Learned from High Latitudes." Journal of Unmanned Vehicle Systems. https://doi.org/10.1139/juvs-2018-0018.

Baena, Susana, Justin Moat, Oliver Whaley, and Doreen S. Boyd. 2017. "Identifying Species from the Air: UAVs and the Very High Resolution Challenge for Plant Conservation." PLOS ONE 12 (11): e0188714. https://doi.org/10.1371/journal.pone.0188714.

Baston, Daniel. 2023. Exactextractr: Fast Extraction from Raster Datasets Using Polygons. Manual. https://CRAN.R-project.org/package=exactextractr.

Berg, J. H. van den, and Elzie Catharina. 2013. Detection, Quantification and Monitoring of Prosopis in the Northern Cape Province of South Africa Using Remote Sensing and GIS. https://doi.org/10.4314/SAJG.V2I2.

Campos, Claudia M., Valeria E. Campos, Arnaldo Mongeaud, Carlos E. Borghi, Claudia De Los Ríos, and Stella M. Giannoni. 2011. "Relationships between Prosopis Flexuosa (Fabaceae) and Cattle in the Monte Desert: Seeds, Seedlings and Saplings on Cattle-Use Site Classes." Revista Chilena de Historia Natural 84 (2): 289–299. https://doi.org/10.4067/S0716-078X2011000200013.

Chen, Ang, Cong Xu, Min Zhang, Jian Guo, Xiaoyu Xing, Dong Yang, Bin Xu, and Xiuchun Yang. 2024. "Cross-Scale Mapping of above-Ground Biomass and Shrub Dominance by Integrating UAV and Satellite Data in Temperate Grassland." Remote Sensing of Environment 304 (April): 114024. https://doi.org/10.1016/j.rse.2024.114024.

Chilume, Tshegofatso, Kashe Keotshepile, and Cate Macinnis-Ng. 2025. "The Woody Invasive Plant Neltuma Juliflora Influences Plant Species Richness and Soil Chemistry in Botswana." Journal of Vegetation Science 36 (5): e70069. https://doi.org/10.1111/jvs.70069.

Cunliffe, Andrew M., Richard Brazier, and Karen Anderson. 2016. "Drone-Acquired Structure-from- Motion Photogrammetry for High-Precision Measurements of Biomass in Semi-Arid Rangelands." Geophysical Research Abstracts 18 (EGU2016-7888).

Dougill, Andrew, Lawrence Akanyang, Jeremy Perkins, Frank Eckardt, Lindsay Stringer, Nicola Favretto, Julius Atlhopheng, and Kutlwano Mulale. 2016. "Land Use, Rangeland Degradation and Ecological Changes in the Southern Kalahari, Botswana." African Journal of Ecology 54 (March): 59–67. https://doi.org/10.1111/aje.12265.

Eckert, Sandra, Purity R. Mbaabu, Wai-Tim Ng, and Albrecht Ehrensperger. 2025. "Distribution and Spread of Prosopis Juliflora in Eastern Africa." In The Ecology and Management of Invasive Prosopis Trees in Eastern Africa, 53–71. CABI Invasives Series. https://doi.org/10.1079/9781800623644.0004.

Eschen, René, Omega Emmanuel Kaaya, Charles Joseph Kilawe, Barnabas Philip Malila, John Richard Mbwambo, Mickfanaka Steven Mwihomeke, and Winnie Nunda. 2024. "Adoption of a Sustainable Land Management Practice for Invasive Prosopis Juliflora in East Africa." CABI Agriculture and Bioscience 5 (1): 113. https://doi.org/10.1186/s43170-024-00315-1.

Frazier, Amy E., and Le Wang. 2011. "Characterizing Spatial Patterns of Invasive Species Using Sub-Pixel Classifications." Remote Sensing of Environment 115 (8): 1997–2007. https://doi.org/10.1016/j.rse.2011.04.002.

Habibi, Luthfan Nur, Tsutomu Matsui, and Takashi S. T. Tanaka. 2024. "Critical Evaluation of the Effects of a Cross-Validation Strategy and Machine Learning Optimization on the Prediction Accuracy and Transferability of a Soybean Yield Prediction Model Using UAV-Based Remote Sensing." Journal of Agriculture and Food Research 16 (June): 101096. https://doi.org/10.1016/j.jafr.2024.101096.

Heermans, Ben, Jacques van Rooyen, Richard Fynn, Duan Biggs, Matthew Lewis, and John McNutt. 2021. "Husbandry and Herding: A Community-Based Approach to Addressing Illegal Wildlife Trade in Northern Botswana." Frontiers in Conservation Science 2 (September). https://doi.org/10.3389/fcosc.2021.675493.

Hijmans, Robert J. 2023. Terra: Spatial Data Analysis. Manual. https://rspatial.org/.

Hughes, Colin E., Jens J. Ringelberg, Gwilym P. Lewis, and Santiago A. Catalano. 2022. "Disintegration of the Genus Prosopis L. (Leguminosae, Caesalpinioideae, Mimosoid Clade)." PhytoKeys 205 (August): 147–189. https://doi.org/10.3897/phytokeys.205.75379.

Hussain, M. Iftikhar, Ross Shackleton, Ali El-Keblawy, Luís González, and M. Mar Trigo. 2021. "Impact of the Invasive Prosopis Juliflora on Terrestrial Ecosystems." In Sustainable Agriculture Reviews 52, edited by Eric Lichtfouse, 223–278. Cham: Springer International Publishing. https://doi.org/10.1007/978-3-030-73245-5_7.

Integrated Mesquite (Prosopis Species) Management Strategy for Botswana. 2022. Botswana Ministry of Environment and Tourism.

Kabelo, T. I., E. M. Fana, J. M. Hyera, and K. Lebani. 2023. "A Review of Foot-and-Mouth Disease Status and Control Measures in Botswana." Tropical Animal Health and Production 55 (4): 278. https://doi.org/10.1007/s11250-023-03674-5.

Kamiri, Hellen Wangechi, Simon K. Choge, and Mathias Becker. 2024. "Management Strategies of Prosopis Juliflora in Eastern Africa: What Works Where?" Diversity 16 (4): 251. https://doi.org/10.3390/d16040251.

Kattenborn, Teja, Javier Lopatin, Michael Förster, Andreas Christian Braun, and Fabian Ewald Fassnacht. 2019. "UAV Data as Alternative to Field Sampling to Map Woody Invasive Species Based on Combined Sentinel-1 and Sentinel-2 Data." Remote Sensing of Environment 227 (June): 61–73. https://doi.org/10.1016/j.rse.2019.03.025.

Kebede, Almaz Tadesse, and D. Layne Coppock. 2015. "Livestock-Mediated Dispersal of Prosopis Juliflora Imperils Grasslands and the Endangered Grevy's Zebra in Northeastern Ethiopia." Rangeland Ecology & Management 68 (5): 402–407. https://doi.org/10.1016/j.rama.2015.07.002.

Keeping, Derek, Njoxlau Kashe, Horekwe (Karoha) Langwane, Panana Sebati, Nicholas Molese, Marie- Charlotte Gielen, Amo Keitsile-Barungwi, Quashe (/Uase) Xhukwe, and !Nate (Shortie) Brahman. 2019. "Botswana's Wildlife Losing Ground as Kalahari Wildlife Management Areas (WMAs) Are Dezoned for Livestock Expansion." Preprint, bioRxiv, March 14. https://doi.org/10.1101/576496.

Kgosikoma, Olaotswe Ernest, and Nnyaladzi Batisani. 2014. "Livestock Population Dynamics and Pastoral Communities' Adaptation to Rainfall Variability in Communal Lands of Kgalagadi South, Botswana." Pastoralism 4 (December). https://doi.org/10.1186/s13570-014-0019-0.

Lang, Michel, Martin Binder, Jakob Richter, Patrick Schratz, Florian Pfisterer, Stefan Coors, Quay Au, Giuseppe Casalicchio, Lars Kotthoff, and Bernd Bischl. 2019. "mlr3: A Modern Object-Oriented Machine Learning Framework in R." Journal of Open Source Software. https://doi.org/10.21105/joss.01903.

Leutner, Benjamin, Ned Horning, and Jakob Schwalb-Willmann. 2022. RStoolbox: Tools for Remote Sensing Data Analysis. Manual. https://CRAN.R-project.org/package=RStoolbox.

Linnenbrink, Jan, Carles Milà, Marvin Ludwig, and Hanna Meyer. 2024. "kNNDM CV: k-Fold Nearest-Neighbour Distance Matching Cross-Validation for Map Accuracy Estimation." Geoscientific Model Development 17: 5897–5912. https://doi.org/10.5194/gmd-17-5897-2024.

Lovelace, Robin, Jakub Nowosad, and Jannes Muenchow. 2019. Geocomputation with R. https://r.geocompx.org/.

Lyons, Mitchell B., David A. Keith, Stuart R. Phinn, Tanya J. Mason, and Jane Elith. 2018. "A Comparison of Resampling Methods for Remote Sensing Classification and Accuracy Assessment." Remote Sensing of Environment 208 (April): 145–153. https://doi.org/10.1016/j.rse.2018.02.026.

Maimaitijiang, Maitiniyazi, Vasit Sagan, Paheding Sidike, Ahmad M. Daloye, Hasanjan Erkbol, Felix B. Fritschi, et al. 2020. "Crop Monitoring Using Satellite/UAV Data Fusion and Machine Learning." Remote Sensing 12 (9). https://doi.org/10.3390/rs12091357.

Malongweni, Siviwe Odwa, Kaya Mrubata, Johan van Tol, Mohamed A. M. Abd Elbasit, and Douglas M. Harebottle. 2025. "Consequences of Invasive Prosopis (Mesquite) on Vegetation, Soil Health, Biodiversity, and Compliance of Management Practices in South African Rangelands: A Review." Grasses 4 (1). https://doi.org/10.3390/grasses4010002.

McMillan, Nicholas A., Samuel D. Fuhlendorf, Craig A. Davis, Robert G. Hamilton, Landon. K. Neumann, and Samantha M. Cady. 2023. "A Plea for Scale, and Why It Matters for Invasive Species Management, Biodiversity and Conservation." Journal of Applied Ecology 60 (7): 1468–1480. https://doi.org/10.1111/1365-2664.14411.

Meroni, Michele, Wai-Tim Ng, Felix Rembold, Ugo Leonardi, Clement Atzberger, Hussein Gadain, and Muse Shaiye. 2017. "Mapping Prosopis Juliflora in West Somaliland with Landsat 8 Satellite Imagery and Ground Information." Land Degradation & Development 28 (2): 494–506. https://doi.org/10.1002/ldr.2611.

Meyer, Hanna, and Edzer Pebesma. 2021. "Predicting into Unknown Space? Estimating the Area of Applicability of Spatial Prediction Models." Methods in Ecology and Evolution 12 (9): 1620–1633. https://doi.org/10.1111/2041-210X.13650.

Meyer, Hanna, and Edzer Pebesma. 2022. "Machine Learning-Based Global Maps of Ecological Variables and the Challenge of Assessing Them." Nature Communications 13 (1): 2208. https://doi.org/10.1038/s41467-022-29838-9.

Meyer, Hanna, Carles Milà, Marvin Ludwig, Jan Linnenbrink, and Fabian Schumacher. 2024. CAST: 'caret' Applications for Spatial-Temporal Models. R package. https://CRAN.R-project.org/package=CAST.

Milà, Carles, Jorge Mateu, Edzer Pebesma, and Hanna Meyer. 2022. "Nearest Neighbour Distance Matching Leave-One-Out Cross-Validation for Map Validation." Methods in Ecology and Evolution 13 (6): 1304–1316. https://doi.org/10.1111/2041-210X.13851.

Mosweu, Samuel, Christopher Munyati, Tibangayuka Kabanda, Moffat Setshogo, and Mbaki Muzila. 2013. "Prosopis L. Invasion in the South-Western Region of Botswana: The Perceptions of Rural Communities and Management Options." Natural Resources 4 (8): 8. https://doi.org/10.4236/nr.2013.48061.

Müllerová, Jana, Giuseppe Brundu, André Große-Stoltenberg, Teja Kattenborn, and David M. Richardson. 2023. "Pattern to Process, Research to Practice: Remote Sensing of Plant Invasions." Biological Invasions. https://doi.org/10.1007/s10530-023-03150-z.

Muzila, Mbaki, Moffat Setshogo, Baleseng Moseki, and Rachel Morapedi. 2010. "An Assessment of Prosopis L. in the Bokspits Area, South-Western Botswana, Based on Morphology." AJPSB 5 (November): 75–80.

Ndhlovu, Thabisisani, Suzanne J Milton, and Karen J Esler. 2016. "Impact of Prosopis (Mesquite) Invasion and Clearing on Vegetation Species Composition and Diversity in Semi-Arid Nama-Karoo Rangeland, South Africa." African Journal of Range & Forage Science 33 (2): 101–110. https://doi.org/10.2989/10220119.2016.1184181.

Ogungbuyi, Michael Gbenga, Caroline Mohammed, Andrew M. Fischer, Darren Turner, Jason Whitehead, Matthew Tom Harrison, et al. 2024. "Integration of Drone and Satellite Imagery Improves Agricultural Management Agility." Remote Sensing 16 (24). https://doi.org/10.3390/rs16244688.

Page, Michael T., Humberto L. Perotto-Baldivieso, J. Alfonso Ortega-S, Evan P. Tanner, Jay P. Angerer, Rider C. Combs, Annalysa M. Camacho, et al. 2022. "Evaluating Mesquite Distribution Using Unpiloted Aerial Vehicles and Satellite Imagery." Rangeland Ecology & Management 83 (July): 91–101. https://doi.org/10.1016/j.rama.2022.03.007.

Pebesma, Edzer. 2018. "Simple Features for R: Standardized Support for Spatial Vector Data." The R Journal 10 (1): 439–446. https://doi.org/10.32614/RJ-2018-009.

Pebesma, Edzer, and Roger Bivand. 2023. Spatial Data Science: With Applications in R. Chapman and Hall/CRC. https://doi.org/10.1201/9780429459016.

Pix4Dmapper. 2022. V. 4.8.0. Pix4D.

Poel, P van der, and J Timberlake. 1980. Sand Dune Stabilisation Bokspits Botswana. Division of Land Utilisation, Department of Agricultural Field Services, Ministry of Agriculture, Botswana.

Radočaj, Dorijan, Ivan Plaščak, and Mladen Jurišić. 2025. "A Comparative Assessment of Regular and Spatial Cross-Validation in Subfield Machine Learning Prediction of Maize Yield from Sentinel-2 Phenology." Eng 6 (10): 270. https://doi.org/10.3390/eng6100270.

Rajah, Perushan, John Odindi, Onisimo Mutanga, and Zolo Kiala. 2019. "The Utility of Sentinel-2 Vegetation Indices (VIs) and Sentinel-1 Synthetic Aperture Radar (SAR) for Invasive Alien Species Detection and Mapping." Nature Conservation 35 (June): 41–61. https://doi.org/10.3897/natureconservation.35.29588.

Rebelo, Alanna J., Shaeden Gokool, Petra B. Holden, and Mark G. New. 2021. "Can Sentinel-2 Be Used to Detect Invasive Alien Trees and Shrubs in Savanna and Grassland Biomes?" Remote Sensing Applications: Society and Environment 23 (August): 100600. https://doi.org/10.1016/j.rsase.2021.100600.

Robinson, T. P., G. W. Wardell-Johnson, G. Pracilio, C. Brown, R. Corner, and R. D. van Klinken. 2016. "Testing the Discrimination and Detection Limits of WorldView-2 Imagery on a Challenging Invasive Plant Target." International Journal of Applied Earth Observation and Geoinformation 44 (February): 23–30. https://doi.org/10.1016/j.jag.2015.07.004.

Rufin, Philippe, David Frantz, Lin Yan, and Patrick Hostert. 2021. "Operational Coregistration of the Sentinel-2A/B Image Archive Using Multitemporal Landsat Spectral Averages." IEEE Geoscience and Remote Sensing Letters 18 (4): 712–716. https://doi.org/10.1109/LGRS.2020.2982245.

Schachtschneider, Klaudia, and Edmund C. February. 2013. "Impact of Prosopis Invasion on a Keystone Tree Species in the Kalahari Desert." Plant Ecology 214 (4): 597–605.

Shackleton, Ross T., David C. Le Maitre, Brian W. Van Wilgen, and David M. Richardson. 2015. "The Impact of Invasive Alien Prosopis Species (Mesquite) on Native Plants in Different Environments in South Africa." South African Journal of Botany 97 (March): 25–31. https://doi.org/10.1016/j.sajb.2014.12.008.

Singh, Kunwar K., Thilina D. Surasinghe, and Amy E. Frazier. 2024. "Systematic Review and Best Practices for Drone Remote Sensing of Invasive Plants." Methods in Ecology and Evolution 15 (6): 998–1015. https://doi.org/10.1111/2041-210X.14330.

Slade, Glenn, Dominic Fawcett, Andrew M. Cunliffe, Richard E. Brazier, Kamal Nyaupane, Marguerite Mauritz, Sergio Vargas, and Karen Anderson. 2023. "Optical Reflectance across Spatial Scales—an Intercomparison of Transect-Based Hyperspectral, Drone, and Satellite Reflectance Data for Dry Season Rangeland." Drone Systems and Applications 11 (January): 1–20. https://doi.org/10.1139/dsa-2023-0003.

Smith, Geoffrey M., and Edward J. Milton. 1999. "The Use of the Empirical Line Method to Calibrate Remotely Sensed Data to Reflectance." International Journal of Remote Sensing 20 (13): 2653–2662. https://doi.org/10.1080/014311699211994.

Statistics Botswana. 2011. Kgalagadi South Sub District. Population and Housing Census Selected Indicators for Villages and Localities. Volume 11.0. Statistics Botswana.

Van Rooyen, Noel, and Gretel Van Rooyen. 2019. Flowering Plants of the Southern Kalahari. Ekotrust.
